``Interestingness'' in Mathematical Theory Exploration

### 2019-10-04 PhDThesis/bucketing.tex
Incorporate feedback

Despite the clever search strategies employed by theory exploration tools, they
all rely on enumerating combinations of definitions in some way, which can cause
infeasible running times on larger inputs (based on experiments with our Theory
Exploration Benchmark,%in \S~\ref{sec:benchmark}
this happens above a few dozen
definitions). To avoid this, users of these systems must carefully select only
small subsets of their definitions to explore at a time.

@@

In this chapter we will:

@@

@@

\begin{figure}
  \centering
  \scalebox{0.45}{\input{images/steppedall.pgf}}
  \scalebox{0.45}{\input{images/steppednontoxic.pgf}}
  \caption{Kaplan-Meier survival plot for running \quickspec{} on inputs
    containing various numbers of definitions, sampled from TIP. The x axis
    denotes running time, which we cut short after 300 seconds. The height of
    each line shows the proportion of \quickspec{} runs which were still going at
    that time (lower is better). First plot is for all TIP definitions, second
    plot removes runs given ``toxic'' definitions.}
  \label{fig:survival}
\end{figure}

@@

\begin{figure}
  \scalebox{0.45}{\input{images/timeoutsall.pgf}}
  \scalebox{0.45}{\input{images/timeoutsnontoxic.pgf}}
  \caption{Proportion of samples which timed out per size, with least-squares
    linear regression. First plot is for all TIP definitions, second removes
    runs given ``toxic'' definitions.}
  \label{fig:tailsize}
\end{figure}

@@

\begin{figure}
  \scalebox{0.45}{\input{images/proportionsall.pgf}}
  \scalebox{0.45}{\input{images/proportionsnontoxic.pgf}}
  \caption{Definitions, ordered by the ratio of successes to failures of the
    runs they appeared in. First graph contains all TIP definitions, showing
    ``toxic'' definitions which always failed. Second graph only contains runs
    without any toxic definitions.}
  \label{fig:proportions}
\end{figure}

@@

To investigate the plausibility of such definitions, the ratio of successful and
timed-out runs is plotted for each name in Figure~\ref{fig:proportions}; showing
that five definitions appeared \emph{only} in failing inputs, and seem likely to
be ``toxic''. These definitions are named \texttt{mul3}, \texttt{mul3acc},
\texttt{mult2}, \texttt{op} and \texttt{qexp}; their definitions appear in
Figure~\ref{fig:faildefs}. All of these are functions of Peano-encoded natural
numbers (\texttt{Nat}), and they cause exploration to time out by either
exhausting the RAM with too many expressions to explore, % TODO{2019-04-27} CAN QSPEC2 AVOID THIS?
or by generating such deeply-nested outputs that comparing them takes too long.
% TODO{2019-04-27} CAN WE PUT TIME AND MEMORY LIMITS ON EACH ATTEMPTED EVALUATION?

@@

\begin{itemize}
\item Avoid ``toxic'' definitions: those producing exponentially large outputs,
  and/or taking many arguments (leading to a large number of combinations). In
  particular, tail-recursive functions which expose their accumulator argument
  are harder to explore.
\item Smaller signatures are faster to explore, so should be preferred if
  possible.
\end{itemize}

@@

Our final criterion concerns the ``quality'' of the selected subsets. This
depends crucially on how we define whether a property is ``interesting''. We can
use the same ground truth definition as our Theory Exploration Benchmark: a
property is ``interesting'' if it appears in a particular corpus (in our case
TIP~\cite{claessen2015tip}), and ``uninteresting'' if not. Whilst na\"ive, this
approach allows the criterion to be quantified: the quality of the selected
subsets is the proportion of the whole signature's interesting properties that
also apply to the subsets taken individually.

@@

\begin{figure}
  % \input{images/bounds.pgf}
  \includegraphics[width=\textwidth]{images/bounds.png}
  \caption{The average quality (proportion of ground-truth properties available)
    when splitting signatures sampled using the Theory Exploration Benchmark
    into subsets. Each plot shows division into a different number of subsets.
    The shaded area is bounded by the optimal and pessimal methods, which are
    can only produce $N$ subsets for signatures of size $\geq N$. Grey lines
    show pseudorandom selection, for comparison; these extend into smaller
    signature sizes since empty subsets are permitted.}
  \label{fig:bounds}
\end{figure}

@@

The results of this method are shown in Figure~\ref{fig:bucketing}, along with
the pseudorandom method as a control. The results of clustering turn out to be
no better than random; in fact the curves show surprisingly similar patterns.

>>>>

Despite the clever search strategies employed by theory exploration tools, they
all rely on enumerating combinations of definitions in some way, which can cause
infeasible running times for input signatures with more than around a dozen
definitions (based on experiments with our Theory Exploration Benchmark, in
\S~\ref{sec:benchmark}. To avoid this, users of these systems must carefully
select only small subsets of their definitions to explore at a time.

@@

\iffalse TODO: Alison: This is very clear and good. Each chapter should have
this at the start \fi
In this chapter we will:

@@

\iffalse
TODO: Alison: Are you aiming to produce a set of recommendation for TE
researchers to follow here?

My thoughts: Not really. We are identifying an open problem (our attempt at a
solution turns out not to work :( )
\fi

@@

\begin{figure}
  \centering
  \scalebox{0.5}{\input{images/steppedall.pgf}}
  \scalebox{0.5}{\input{images/timeoutsall.pgf}}
  \scalebox{0.5}{\input{images/steppednontoxic.pgf}}
  \scalebox{0.5}{\input{images/timeoutsnontoxic.pgf}}
  \caption{Kaplan-Meier survival plot for running \quickspec{} on inputs
    containing various numbers of definitions, sampled from TIP. The x axis
    denotes running time, which we cut short after 300 seconds. The height of
    each line shows the proportion of \quickspec{} runs which were still going
    at that time (lower is better). First plot is for all TIP definitions,
    second plot removes runs given ``toxic'' definitions.
    \iffalse
    TODO: Alison: The key needs further explanation (either here or in the caption)
    TODO: Alison: Label the y axis (on both)
    \fi}
  \label{fig:survival}
\end{figure}

@@

\iffalse
\begin{figure}
  \scalebox{0.45}{\input{images/timeoutsall.pgf}}
  \scalebox{0.45}{\input{images/timeoutsnontoxic.pgf}}
  \caption{Proportion of samples which timed out per size, with least-squares
    linear regression. First plot is for all TIP definitions, second removes
    runs given ``toxic'' definitions.}
  \label{fig:tailsize}
\end{figure}
\fi

@@

\begin{figure}
  \scalebox{0.45}{\input{images/proportionsall.pgf}}
  \scalebox{0.45}{\input{images/proportionsnontoxic.pgf}}
  \caption{Definitions, ordered by the ratio of successes to failures of the
    runs they appeared in. The first graph contains all TIP definitions, showing
    ``toxic'' definitions which always failed. The second graph only contains
    runs without any toxic definitions.
    \iffalse
    TODO: Alison: ? (ME: The green/red bar graphs appear as empty white boxes in
    the PDF Alison sent over. Check how it's rendered!)
    TODO: The x-axis ticks are crap (all overlapping); turn them off, put a
    single label like ``Definitions (ordered by failure rate)''
    \fi}
  \label{fig:proportions}
\end{figure}

@@

To investigate the plausibility of such definitions, we plotted the ratio of
successful and timed-out runs for each name in Figure~\ref{fig:proportions}; showing
that five definitions appeared \emph{only} in failing inputs, and seem likely to
be ``toxic''. These definitions are named \texttt{mul3}, \texttt{mul3acc},
\texttt{mult2}, \texttt{op} and \texttt{qexp}; their definitions appear in
Figure~\ref{fig:faildefs}. All of these are functions of Peano-encoded natural
numbers (\texttt{Nat}), and they cause exploration to time out by either
exhausting the RAM with too many expressions to explore, % TODO{2019-04-27} CAN QSPEC2 AVOID THIS?
or by generating such deeply-nested outputs that comparing them takes too long.
% TODO{2019-04-27} CAN WE PUT TIME AND MEMORY LIMITS ON EACH ATTEMPTED EVALUATION?

@@

\begin{itemize}
\item Avoid ``toxic'' definitions: those producing exponentially large outputs,
  and/or taking many arguments (leading to a large number of combinations)
  \iffalse TODO: Alison: Is it always obvious which are toxic? \fi. In
  particular, tail-recursive functions which expose their accumulator argument
  are harder to explore. \iffalse TODO: Alison: These should be in the
  contributions and further recommendations based on your other findings in this
  chapter \fi
\item Smaller signatures are faster to explore, so should be preferred if
  possible.
\end{itemize}

@@

Our final criterion \iffalse TODO: Alison: This looks odd since you haven't
mentioned criteria before \fi concerns the ``quality'' of the selected
subsets. This depends crucially on how we define whether a property is
``interesting''. We can use the same ground truth definition as our Theory
Exploration Benchmark: a property is ``interesting'' if it appears in a
particular corpus (in our case TIP~\cite{claessen2015tip}), and
``uninteresting'' if not. Whilst na\"ive, this approach allows the criterion to
be quantified: the quality of the selected subsets is the proportion of the
whole signature's interesting properties that also apply to the subsets taken
individually.

@@

\begin{figure}
  % \input{images/bounds.pgf}
  \includegraphics[width=\textwidth]{images/bounds.png}
  \caption{The average quality (proportion of ground-truth properties available)
    when splitting signatures sampled using the Theory Exploration Benchmark
    into subsets. Each plot shows division into a different number of subsets.
    The shaded area is bounded by the optimal and pessimal methods, which can
    only produce $N$ subsets for signatures of size $\geq N$. Grey lines show
    pseudorandom selection, for comparison; these extend into smaller signature
    sizes since empty subsets are permitted.}
  \label{fig:bounds}
\end{figure}

@@

The results of this method are shown in Figure~\ref{fig:bucketing}, along with
the pseudorandom method as a control. The results of clustering turn out to be
no better than random; in fact the curves show surprisingly similar patterns.
\iffalse
TODO: Alison: Does this translate into a recommendation (e.g. not to
use clustering)? There should be some discussion of this - if not here then
point to another chapter where you discuss these results

TODO: Alison: Needs a summary section
\fi
